---
title: "Generalized Linear Mixed Models"
format: 
  pdf:
    code-line-numbers: true
---

```{r}
#| message: false
#| warning: false
#| echo: false
# load libraries
library(tidyverse)
library(dotwhisker)
library(effects)
library(car)
library(splines)
library(performance)
library(DHARMa)
library(ggplot2)
library(gridExtra)
library(lme4)
library(glmmTMB)
library(HSAUR3)
library(GLMMadaptive)
library(MASS)
library(broom.mixed)
library(dplyr)
library(brms)

```

## Study 1:

## a.

For this analysis, I have chosen 'total_medals' as the response variable. The fixed-effect predictor variables are log(gdp), log(pop), and year_centered (which is derived by subtracting the minimum year from each year to center the data around zero). The random effects include both random intercepts and slopes for team, and I will also account for random intercepts and slopes across year_centered. The maximal model specification would look like this:\

```         
full_model <- lmer(total_medals ~ log(gdp) + log(pop) + year_centered + 
                   (1 + log(gdp) + log(pop) + year_centered | team) + 
                   (1 + log(gdp) + log(pop) | year_centered), 
                   data = olympic_trans)
```

## b.

In our dataset, there is only one observation per country-year combination, which means we do not have sufficient data to support the inclusion of random effects for both team and year_centered within each group.

The response variable total_medals has a highly right_skewed distribution, with most of the data concentrated around lower medal counts and a long tail extending towards higher counts. Poisson distribution is more appropriate for modeling this data. I will use glmmTMB function instead lmer to built the model.

```{r}
#| message: false


# url to olympic dataset 
url <- "https://raw.githubusercontent.com/bbolker/stats720/main/data/olymp1.csv"

# Import the dataset
olympic_data <- read_csv(url)

```

```{r}
summary(olympic_data)
```

```{r}
missing_values <- sapply(olympic_data, function(x) sum(is.na(x)))
missing_values
```

```{r}
# remove all missing values
olympic_clean <- na.omit(olympic_data)
sapply(olympic_clean,function(x) sum(is.na(x)))
```

```{r}
hist (olympic_clean$n, main='Histogram of Medal Counts', xlab = 'Number of Medals', col = 'skyblue', breaks = 30)

boxplot(olympic_clean$n, main='Box Plot of Medal Counts', ylab='Number of Medals')

```

```{r}
# choose the total number of variable 
# transform the column 'n' and 'medal' to 'Gold'/'Silver'/'Bronze'
olympic_trans <- olympic_clean %>%
  pivot_wider(names_from = medal, values_from = n, values_fill = list(n = 0)) %>%
  group_by(team, year, gdp, pop) %>%
  summarize(
    gold = sum(Gold, na.rm = TRUE),
    silver = sum(Silver, na.rm = TRUE),
    bronze = sum(Bronze, na.rm = TRUE)
  )
```

```{r}

olympic_trans$total_medals = olympic_trans$gold+olympic_trans$silver+olympic_trans$bronze
olympic_trans$year_centered = olympic_trans$year - min(olympic_trans$year) 
olympic_trans=olympic_trans[,!(names(olympic_trans) %in% c('gold','silver','bronze'))]
head(olympic_trans)
```

```{r}
full_model <- glmer(total_medals ~ 1 + log(gdp) + log(pop) + year_centered + 
                        (1 + log(gdp) + log(pop) + year_centered | team) + 
                        (1 + log(gdp) + log(pop) | year_centered), 
                      family = poisson, 
                      data = olympic_trans)
```

## c.

I initially started with the full model, but it encountered convergence issues. First, I removed the random effect term for `year`. Second, I simplified the random slope structure. I ultimately arrived at the `model_slop` model, using the `glmmTMB` function with a negative binomial link function to handle overdispersion. These adjustments resolved the convergence problem.

```{r}
model_slop <- glmmTMB(total_medals ~ 1 + log(gdp) + log(pop) + year_centered + 
                        (1 +  year_centered | team),
                      family = poisson, 
                      data = olympic_trans)
```

```{r}
check_overdispersion(model_slop)

```

```{r}
model_slop_NB <- glmmTMB(
  total_medals ~ 1 + log(gdp) + log(pop) + year_centered + 
    (1 +  year_centered | team),
  family = nbinom2(link = "log"), 
  data = olympic_trans)
```

```{r}
check_overdispersion(model_slop_NB)

```

## d. exploratory plots

```{r}

# Plot with geom_line for individual groups
ggplot(olympic_trans, aes(x = year_centered, y = total_medals, group = team, color = team)) +
  geom_line(alpha = 0.6) +
  labs(
    title = "Total Medals Over Time by Team",
    x = "Year (Centered)",
    y = "Total Medals"
  ) +
  theme_minimal() +
  theme(legend.position = "none")  # Remove legend if there are many groups
```

```{r}
# Total_medals vs Population group by team
ggplot(olympic_trans, aes(x = pop, y = total_medals, group = team)) +
  geom_path(aes(color = year), alpha = 0.5) +
  geom_point(aes(color = year), size = 2) +
  scale_color_viridis_c() +
  labs(x = "Population", y = "Total Medals") +
  theme_minimal() +
  theme(legend.position = "right")
```

```{r}
# Total_medals vs GDP group by team
ggplot(olympic_trans, aes(x = gdp, y = total_medals, group = team)) +
  geom_path(aes(color = year), alpha = 0.5) +
  geom_point(aes(color = year), size = 2) +
  scale_color_viridis_c() +
  labs(x = "GDP", y = "Total Medals") +
  theme_minimal() +
  theme(legend.position = "right")
```

## e. fitting models

I built four models by successively incorporating splines, zero-inflation, rank reduction, and interaction terms. I then compared these models using diagnostic plots and AIC values. Despite adding interaction terms and splines, all the models continue to show significant dispersion and structural patterns in their residuals. However, when I use `check_overdispersion()` on these models, it indicates that there is no overdispersion. I’m unsure why the diagnostic plots suggest overdispersion even though `check_overdispersion()` does not detect it.

```{r}
# adding spline 
model_slop_NB_adjusted_1 <- glmmTMB(
  total_medals ~ 1 + log(gdp) + log(pop) + ns(year_centered, df=4) + 
    (1 | team),
  family = nbinom2(link = "log"),
  REML = TRUE,
  data = olympic_trans)

```

```{r}
# add zero-inflation
model_slop_NB_adjusted_2 <- glmmTMB(
  total_medals ~ 1 + log(gdp) + log(pop) + ns(year_centered, df=4) + 
    (1 | team),
  family = nbinom2(link = "log"),
  ziformula = ~1,
  REML = TRUE,
  data = olympic_trans)
```

```{r}
# add diag
model_slop_NB_adjusted_3 <- glmmTMB(
  total_medals ~ 1 + log(gdp) + log(pop) + ns(year_centered, df=4) + 
    diag(year_centered|team),
  family = nbinom2(link = "log"),
  ziformula = ~1,
  REML = TRUE,
  data = olympic_trans)
```

```{r}
# add intercection 
model_slop_NB_adjusted_4 <- glmmTMB(
  total_medals ~ 1 + log(gdp) + log(pop) + ns(year_centered, df=4) + log(gdp)*log(pop) + 
    diag(year_centered|team),
  family = nbinom2(link = "log"),
  ziformula = ~1,
  REML = TRUE,
  data = olympic_trans)
```

```{r}
all_fits <- ls(pattern = "^model_slop_NB_adjusted_")
all_fit_list <- mget(all_fits)
bbmle::AICtab(all_fit_list)
```

```{r}
dharma_res_1 <- simulateResiduals(fittedModel = model_slop_NB_adjusted_1)
dharma_res_2 <- simulateResiduals(fittedModel = model_slop_NB_adjusted_2)
dharma_res_3 <- simulateResiduals(fittedModel = model_slop_NB_adjusted_3)
dharma_res_4 <- simulateResiduals(fittedModel = model_slop_NB_adjusted_4)

# Set up a multi-plot layout (3 rows, 1 column)
par(mfrow = c(4, 1))

# Plot residual diagnostics for each model
plot(dharma_res_1, main = "Model_1")
plot(dharma_res_2, main = "Model_2")
plot(dharma_res_3, main = "model_3")
plot(dharma_res_4, main = "model_4")


```

```{r}
dwplot(list(model_slop_NB_adjusted_4, model_slop_NB_adjusted_3)) %>%
  relabel_predictors(c(
    "log(gdp)" = "Log(GDP)",
    "log(pop)" = "Log(Population)",
    "ns(year_centered, df = 4)" = "Year (Spline)"
  )) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Coefficient Plot with Custom Labels",
    x = "Estimate",
    y = "Predictors"
  ) +
  theme_minimal()

```

## Study 2:

## a.

The response variable is `outcome`. The fixed-effect predictor variables are `time` and `visit`, while the random-effect grouping variables are `treatment` and `patientID`.

## c.

The model I am going to try to fit initially would be:

full_model_toenail \<- glmer(outcome_binary \~ time + vist + (1 \| patientID) + (1 \| treatment), data = toenail, family = binomial(link = "logit")

\
In this model:

-   The **fixed effects** estimate the impact of `time` and `visit` (on a logarithmic scale) on the binary outcome variable (`outcome_binary`).

-   The **random effects** account for variability across patients (`patientID`) and treatments (`treatment`).

```{r}

data("toenail", package = "HSAUR3")
head(toenail)
```

```{r}

# explore the data
summary(toenail)
colSums(is.na(toenail))
toenail$outcome_binary <- as.numeric(toenail$outcome == "moderate or severe")
str(toenail)

```

## d.

```{r}
# Spaghetti plot for patientID
ggplot(toenail, aes(x = time, y = visit, group = patientID, color = as.factor(patientID))) +
  geom_line(alpha = 0.6) +
  labs(
    title = "Spaghetti Plot: Time Across Visits by Patient",
    x = "Time",
    y = "Visit Number",
    color = "Patient ID"
  ) +
  theme_minimal() +
  theme(legend.position = "none")  # Hide legend if there are many patients

```

```{r}
# Generalized linear model (quasipoisson family for count data)
ggplot(toenail, aes(x = visit, y = time, color = treatment)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "glm", method.args = list(family = "quasipoisson"), se = FALSE) +
  labs(
    title = "GLM Fit: Quasipoisson Regression for Time",
    x = "Visit Number",
    y = "Time",
    color = "Treatment"
  ) +
  theme_minimal()

```

## e.

My initial model exhibited a singular fit. To find a better-fitting alternative, I explored three different models by simplifying the random effects structure, adding an interaction term, and incorporating spline functions.

```{r}
full_model_toenail <- glmer(
  outcome_binary ~ time + visit + 
    (1 | patientID) + (1 | treatment), 
  data = toenail,
  family = binomial(link = "logit"))


```

```{r}
model_toenail_adjuseted_1 <- glmer(
  outcome_binary ~ time + visit + 
    (1 | patientID), 
  data = toenail,
  family = binomial(link = "logit"))
```

```{r}
model_toenail_adjuseted_2 <- glmer(
  outcome_binary ~ time * visit + 
    (1 | patientID), 
  data = toenail,
  family = binomial(link = "logit"))
```

```{r}
model_toenail_adjuseted_3 <- glmer(
  outcome_binary ~ ns(time, 3) + ns(visit, 3) + 
    (1 | patientID), 
  data = toenail,
  family = binomial(link = "logit"))
```

```{r}
all_fits <- ls(pattern = "^model_toenail_adjuseted_")
all_fit_list <- mget(all_fits)
bbmle::AICtab(all_fit_list)
```

```{r}
dharma_res_1 <- simulateResiduals(fittedModel = model_toenail_adjuseted_1)
dharma_res_2 <- simulateResiduals(fittedModel = model_toenail_adjuseted_2)
dharma_res_3 <- simulateResiduals(fittedModel = model_toenail_adjuseted_3)


# Set up a multi-plot layout (3 rows, 1 column)
par(mfrow = c(3, 1))

# Plot residual diagnostics for each model
plot(dharma_res_1, main = "Model_1")
plot(dharma_res_2, main = "Model_2")
plot(dharma_res_3, main = "model_3")

```

```{r}
dwplot(model_toenail_adjuseted_2) %>%
  relabel_predictors(c(
    "time * visit" = "time * visit"
  )) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Coefficient Plot ",
    x = "Estimate",
    y = "Predictors"
  ) +
  theme_minimal()

```

## f.

```{r}
# a.completely pooled analysis(GLM)
model_glm <- glm(
  outcome_binary ~ time * visit,
  data = toenail,
  family = binomial(link = "logit")
)
```

```{r}
# b.penalized Quasi-Likelihood (PQL)
model_pql <- glmmPQL(
  fixed = outcome_binary ~ time * visit,
  random = ~ 1 | patientID,
  family = binomial(link = "logit"),
  data = toenail
)
```

```{r}
# c. Laplace Approximation
model_laplace <- glmer(
  outcome_binary ~ time * visit + (1 | patientID),
  data = toenail,
  family = binomial(link = "logit")
)
```

```{r}
# d. Adaptive Gauss-Hermite Quadrature (10 and 20 Points)
# With 10 quadrature points
model_agq10 <- glmer(
  outcome_binary ~ time * visit + (1 | patientID),
  data = toenail,
  family = binomial(link = "logit"),
  nAGQ = 10
)

# With 20 quadrature points
model_agq20 <- glmer(
  outcome_binary ~ time * visit + (1 | patientID),
  data = toenail,
  family = binomial(link = "logit"),
  nAGQ = 20
)
```

```{r}

# e. Fit a Bayesian model with default priors
model_bayesian <- brm(
  outcome_binary ~ time * visit + (1 | patientID),
  data = toenail,
  family = bernoulli(link = "logit"),
  prior = c(
    prior(normal(0, 10), class = "b"),        # Weakly informative priors
    prior(normal(0, 10), class = "Intercept"),
    prior(exponential(1), class = "sd")       # Prior for random effect sd
  ),
  iter = 2000, chains = 4
)

```

The comparison coefficien plot shows that all the models have similar estimates for time, visit and time:visit.

The Bayesian model produces a significantly different estimate for the intercept parameter compared to other models.

```{r}
#| waning: false 
# coefficients for GLM
glm_results <- tidy(model_glm, conf.int = TRUE) %>%
  mutate(model = "GLM (Pooled)")

# coefficients for PQL
pql_results <- tidy(model_pql, conf.int = TRUE) %>%
  mutate(model = "PQL")

# coefficients for Laplace
laplace_results <- tidy(model_laplace, conf.int = TRUE) %>%
  mutate(model = "Laplace")

# coefficients for AGQ (10 points)
agq10_results <- tidy(model_agq10, conf.int = TRUE) %>%
  mutate(model = "AGQ (10 points)")

# coefficients for AGQ (20 points)
agq20_results <- tidy(model_agq20, conf.int = TRUE) %>%
  mutate(model = "AGQ (20 points)")

# coefficients for Bayesian model
bayesian_results <- as.data.frame(fixef(model_bayesian, summary = TRUE)) %>%
  rownames_to_column("term") %>%
  rename(estimate = Estimate, conf.low = `Q2.5`, conf.high = `Q97.5`) %>%
  mutate(model = "Bayesian")

```

```{r}
# Combine all results into one data frame
combined_results <- bind_rows(
  glm_results,
  pql_results,
  laplace_results,
  agq10_results,
  agq20_results,
  bayesian_results
)

```

```{r}
# Create the coefficient comparison plot
dwplot(combined_results, by_2sd = FALSE, dot_args = list(size = 2), whisker_args = list(size = 1)) +
  theme_minimal() +
  labs(
    title = "Comparison of Fixed-Effect Estimates Across Models",
    x = "Estimate",
    y = "Fixed-Effect Parameters",
    color = "Model"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "top"
  )

```

## Study 3:

## a.

```{r}
simfun <- function(beta, theta, n_t, n_id) {
    id <- rep(1:n_id, each = n_t)
    time <- rep(0:(n_t-1), times = n_id)
    ttt <- rep(rep(c(0,1), each = n_t), length.out = n_id * n_t)
    
    X <- model.matrix(~ 1 + ttt*time)
    
    random_effects <- rnorm(n_id, 0, theta)
    re <- rep(random_effects, each = n_t)
    
    eta <- X %*% beta + re
    
    p <- 1/(1 + exp(-eta))
    y <- rbinom(length(p), 1, p)
    
    data.frame(
        id = factor(id),
        time = time,
        ttt = factor(ttt, labels = c("control", "treatment")),
        y = y
    )
}


set.seed(123)  
beta_test <- c(1, -0.5, -0.1, 0.05)  
theta_test <- 0.5
n_t_test <- 7    
n_id_test <- 10  

test_data <- simfun(beta_test, theta_test, n_t_test, n_id_test)

print(head(test_data))

test_model <- glmer(y ~ ttt*time + (1|id), 
                    family = binomial, 
                    data = test_data)
print(summary(test_model))
```

```{r}
fitfun <- function(data, nAGQ) {
    extract_coef_ci <- function(model, type = "standard") {
        if (type == "standard") {
            coef_table <- summary(model)$coefficients
            if (inherits(model, "glm")) {
                ci <- confint(model)
            } else {
                est <- coef_table[, "Estimate"]
                se <- coef_table[, "Std. Error"]
                ci <- cbind(
                    est - 1.96 * se,
                    est + 1.96 * se
                )
            }
            result <- data.frame(
                estimate = coef_table[, "Estimate"],
                se = coef_table[, "Std. Error"],
                ci_lower = ci[, 1],
                ci_upper = ci[, 2]
            )
        } else if (type == "pql") {
            coef_table <- summary(model)$tTable
            est <- coef_table[, "Value"]
            se <- coef_table[, "Std.Error"]
            result <- data.frame(
                estimate = est,
                se = se,
                ci_lower = est - 1.96 * se,
                ci_upper = est + 1.96 * se
            )
        }
        return(result)
    }
    
    if (nAGQ == -2) {
        fit <- glm(y ~ ttt * time, family = binomial, data = data)
        results <- extract_coef_ci(fit)
        
    } else if (nAGQ == -1) {
        fit <- glmmPQL(y ~ ttt * time, 
                      random = ~ 1 | id,
                      family = binomial, 
                      data = data)
        results <- extract_coef_ci(fit, type = "pql")
        
    } else if (nAGQ >= 1) {
        fit <- glmer(y ~ ttt * time + (1 | id),
                    family = binomial,
                    data = data,
                    nAGQ = nAGQ)
        results <- extract_coef_ci(fit)
    }
    
    return(results)
}

# test different nAGQ values
test_results_glm <- fitfun(test_data, -2)
test_results_pql <- fitfun(test_data, -1)
test_results_laplace <- fitfun(test_data, 1)

cat("GLM (pooled) results:\
")
print(test_results_glm)
cat("\
PQL results:\
")
print(test_results_pql)
cat("\
Laplace (nAGQ=1) results:\
")
print(test_results_laplace)
```

3\.

```{r}
simfun <- function(n_t, n_id, beta, theta) {
  time <- rep(1:n_t, n_id)
  id <- rep(1:n_id, each = n_t)
  treatment <- rbinom(n_id, 1, 0.5) 
  random_effect <- rnorm(n_id, 0, exp(theta))
  response <- beta[1] + beta[2] * treatment + beta[3] * time +
    beta[4] * treatment * time + random_effect[id] +
    rnorm(n_t * n_id, 0, 0.5) 
  data.frame(id, time, treatment, response)
}
```

```{r}
# Fitting function
fitfun <- function(data) {
  fit <- glmmTMB(
    response ~ treatment * time + (1 | id),
    data = data
  )
  broom.mixed::tidy(fit, effects = "fixed", conf.int = TRUE)
}
```

```{r}
# Simulation and evaluation
evaluate <- function(n_t, beta, theta, n_id, n_sim = 100) {
  results <- replicate(n_sim, {
    sim_data <- simfun(n_t, n_id, beta, theta)
    tryCatch(fitfun(sim_data), error = function(e) NULL)
  }, simplify = FALSE)
  
  results <- bind_rows(results, .id = "simulation")
  true_vals <- data.frame(term = c("(Intercept)", "treatment", "time", "treatment:time"), 
                          true = beta)
  results <- left_join(results, true_vals, by = "term")
  results %>%
    filter(term %in% c("treatment", "treatment:time")) %>%
    group_by(term) %>%
    summarize(
      bias = mean(estimate - true, na.rm = TRUE),
      variance = var(estimate, na.rm = TRUE),
      scaled_rmse = sqrt(mean((estimate / true - 1)^2, na.rm = TRUE)),
      coverage = mean(conf.low <= true & conf.high >= true, na.rm = TRUE)
    )
}
```

```{r}
# Run simulations for n_t = 5 and n_t = 10
set.seed(1234)
beta <- c(-0.6, 0, -0.2, -0.05)
theta <- log(0.2)
n_id <- 300

results_5 <- evaluate(n_t = 5, beta = beta, theta = theta, n_id = n_id)
results_10 <- evaluate(n_t = 10, beta = beta, theta = theta, n_id = n_id)

# Combine results
results <- bind_rows(
  results_5 %>% mutate(n_t = 5),
  results_10 %>% mutate(n_t = 10)
)
```

```{r}
# Visualize results
ggplot(results, aes(x = term, y = bias, fill = factor(n_t))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Bias for beta[2] and beta[4]", y = "Bias", fill = "Time Points")

ggplot(results, aes(x = term, y = variance, fill = factor(n_t))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Variance for beta[2] and beta[4]", y = "Variance", fill = "Time Points")

ggplot(results, aes(x = term, y = scaled_rmse, fill = factor(n_t))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Scaled RMSE for beta[2] and beta[4]", y = "Scaled RMSE", fill = "Time Points")

ggplot(results, aes(x = term, y = coverage, fill = factor(n_t))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Coverage for beta[2] and beta[4]", y = "Coverage", fill = "Time Points")
```
